---
title: "CaLMR Tutorial: Test the Effect of Chronic Inflammation on Rheumatoid Arthritis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{CaLMR Tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

The tutorial demonstrates the CaLMR workflow using the chronic inflammation – Rheumatoid Arthritis 
analysis from the paper as an example, with CRP, IL-6, IL-8, CCL2, and TNF-$\alpha$ 
as the biomarkers and RA as the outcome. 
The package includes sample data (chromosomes 1–2) for format reference.

We cover four stages:

1. Required input data format
2. GWAS QC and harmonization (preparing your own data)
3. Estimating the between-biomarker correlation matrix
4. Running CaLMR


## Prerequisites

```{r install}
# install.packages("devtools")
devtools::install_github("your-username/CaLMR")
library(CaLMR)
```

The full pipeline additionally requires:

- [PLINK](https://www.cog-genomics.org/plink/) for LD clumping
- A 1000 Genomes reference panel in PLINK binary format (`chr1.bim/.bed/.fam`,
  ..., `chr22.bim/.bed/.fam`)
- LD score files for sample-overlap estimation (e.g., from the
  [LDSC repository](https://github.com/bulik/ldsc), 1000 Genomes Phase 3
  European panel)


## Part 1: Input Data Format

CaLMR requires each GWAS summary statistics file to contain the following
columns:

| Column | Description |
|--------|-------------|
| `SNP`  | SNP identifier (rsID) |
| `CHR`  | Chromosome |
| `BETA` | Effect size estimate |
| `SE`   | Standard error |
| `PVAL` | P-value |
| `A1`   | Effect allele |
| `A2`   | Other allele |
| `NEFF` | Effective sample size |

The package includes sample data (chromosomes 1–2 only) to illustrate the
expected format:

```{r sample-data, eval=TRUE}
library(CaLMR)
data_dir <- system.file("extdata", package = "CaLMR")

# Load one biomarker as an example
crp <- readRDS(file.path(data_dir, "crp.RDS"))
head(crp)
str(crp)
```

The sample data files included are:

- `crp.RDS`, `il6.RDS`, `il8.RDS`, `ccl2.RDS`, `tnfa.RDS` — biomarker GWAS
- `ra.RDS` — outcome GWAS
- `corr_biomarker.RDS` — pre-computed 5×5 between-biomarker correlation matrix

These are subsets (chr 1–2) for format reference only. A full analysis
requires genome-wide data.


## Part 2: GWAS QC and Harmonization

Raw GWAS files often use different column names and allele codings. Before
running CaLMR, each file must be harmonized to the format above and aligned to
a common reference genome. Below is a reference workflow using the RA GWAS
(Okada et al.) as an example.

### Loading a reference genome

```{r load-reference}
library(data.table)

# Load 1000 Genomes Build 37 .bim files as the allele reference
reference_dir <- "/path/to/1KG/GRCh37"
bim_files <- file.path(reference_dir, paste0("chr", 1:22, ".bim"))
genome37 <- rbindlist(lapply(bim_files, function(f) {
  fread(f, header = FALSE,
        col.names = c("CHR", "rsid", "CM", "BP", "A1", "A2"))
}))
```

### Harmonizing a single GWAS file

Take the RA GWAS as an exmple:

```{r qc-example}
# Read the raw GWAS
ra_raw <- fread("RA.GWAS.Okada.txt")

# Rename allele columns
setnames(ra_raw, c("A1", "A2"), c("effect_allele", "other_allele"))

# Merge with reference 
merged <- merge(ra_raw, genome37,
                by.x = "SNP", by.y = "rsid",
                suffixes = c(".gwas", ".ref"))

# Keep only SNPs whose alleles match the reference (either orientation)
matched <- merged[
  (effect_allele == A1 & other_allele == A2) |
  (effect_allele == A2 & other_allele == A1)
]

# Flip effect direction when alleles are swapped relative to the reference
flip_idx <- matched[effect_allele == A1 & other_allele == A2, which = TRUE]
matched[flip_idx, beta := -beta]
matched[flip_idx, c("effect_allele", "other_allele") := .(other_allele, effect_allele)]

# Select and rename to CaLMR standard format
ra_qc <- matched[, .(SNP, CHR = CHR.gwas, BP = BP,
                      A1 = effect_allele, A2 = other_allele,
                      BETA = beta, SE = se, PVAL = p, NEFF = Neff)]

saveRDS(as.data.frame(ra_qc), file = "data/ra.RDS")
```


In this analysis, the five inflammatory GWAS are processed following a similar
procedure.
After this step, you should have one `.RDS` file per trait:

```
data/
  crp.RDS, il6.RDS, il8.RDS, ccl2.RDS, tnfa.RDS, ra.RDS
```


## Part 3: Estimating the Between-Biomarker Correlation Matrix

GWAS studies that share participants produce correlated summary statistics
even when the traits are genetically independent. CaLMR corrects for this
using a correlation matrix estimated via LD score regression.

### Estimating with `corr_cal()`

`corr_cal()` takes file paths to the biomarker GWAS and a directory of LD
score files, and returns a K × K matrix:

```{r corr-cal}
traitvec <- c("crp", "il6", "il8", "ccl2", "tnfa")

biomarker_paths <- c(
  crp = "data/crp.RDS",
  il6 = "data/il6.RDS",
  il8 = "data/il8.RDS",
  ccl2 = "data/ccl2.RDS",
  tnfa = "data/tnfa.RDS"
)

corr_KK <- corr_cal(
  biomarker_paths = biomarker_paths,
  traitvec  = traitvec,
  ld_score_dir = "/path/to/ld_scores"
)
```


### Building the (K+1) × (K+1) matrix for CaLMR

`run_calmr()` requires a (K+1) × (K+1) matrix where the last row/column
corresponds to the outcome. Under the two-sample MR assumption (no sample
overlap between biomarker and outcome GWAS), the outcome row/column is filled
with zeros for off-diagonal entries:

```{r build-corr-mat, eval=TRUE}
K <- nrow(corr_KK)
outcome_name <- "ra"

Corr.mat <- matrix(0, nrow = K + 1, ncol = K + 1)
Corr.mat[1:K, 1:K] <- corr_KK
Corr.mat[K + 1, K + 1] <- 1
colnames(Corr.mat) <- c(traitvec, outcome_name)
rownames(Corr.mat) <- c(traitvec, outcome_name)

print(round(Corr.mat, 3))
```


For this tutorial, we provide the pre-computed correlation matrix from the
paper for your reference:

```{r load-corr, eval=TRUE}
data_dir <- system.file("extdata", package = "CaLMR")
traitvec <- c("crp", "il6", "il8", "ccl2", "tnfa")

corr_KK <- readRDS(file.path(data_dir, "corr_biomarker.RDS"))
print(round(corr_KK, 3))
```


## Part 4: Running CaLMR

### Setup

Define the file paths to QC'd genome-wide GWAS data and the analysis
parameters. Both `corr_cal()` and `run_calmr()` take the same
`biomarker_paths` format — a named character vector of file paths.

```{r setup-analysis}
biomarker_paths <- c(
  crp = "data/crp.RDS",
  il6 = "data/il6.RDS",
  il8 = "data/il8.RDS",
  ccl2 = "data/ccl2.RDS",
  tnfa = "data/tnfa.RDS"
)
outcome_path <- "data/ra.RDS"

# P-value thresholds for IV selection (per biomarker)
pval_threshold <- c(
  crp = 5e-5,
  il6 = 1e-4,
  il8 = 1e-4,
  ccl2 = 1e-4,
  tnfa = 1e-4
)

# specify the pathways
plink_path    <- "/path/to/plink"
reference_dir <- "/path/to/1KG/GRCh37"
```

### Run the CaLMR piprline


The `sign` argument encodes prior knowledge about the direction of each
biomarker's relationship with the latent exposure (1 = positive,
-1 = negative, 0 = unknown). Here, all five biomarkers are expected to
increase with chronic inflammation levels:

```{r run-uni}
result_uni <- run_calmr(
  method = "uni",
  biomarker_paths = biomarker_paths,
  outcome_path = outcome_path,
  outcome_name = "ra",
  traitvec = traitvec,
  sign = c(1, 1, 1, 1, 1),
  Corr.mat = Corr.mat,
  pval_threshold = pval_threshold,
  reference_dir = reference_dir,
  plink_path = plink_path,
  T_iter = 10000,
  burnin = 3000
)

# Main result: p-value, rejection decision, and estimated direction
result_uni$calmr_result

# 95% credible intervals for model parameters
result_uni$CI

```



## Tips

**Choosing p-value thresholds.** The thresholds in `pval_threshold` control
how many SNPs enter the IV selection. A good starting point is 5e-6 for
large GWAS. 

**Minimum SNPs.** If `run_calmr` returns `NULL`, too few SNPs survived
preprocessing. Try relaxing `pval_threshold`, or reducing LD clumping
stringency (`r2`, `kb`).

**Confounder exclusion.** If you know of pleiotropic SNPs, exclude them 
via the `confounder` argument.

**MCMC tuning.** The defaults (`T_iter = 3000`, `burnin = 1500`) are
reasonable for exploratory analyses. For inspection, increase to
`T_iter = 10000, burnin = 3000` and check convergence via
`result$mcmc.detail`.


## Session Info

```{r session-info, eval=TRUE}
sessionInfo()
```
